---
title: Manifest Destiny Report
output: html_document
author: Trevor Day
date: today
toc: true
---

```{r load_data, include = FALSE}

library(tidyverse)
library(patchwork)
library(knitr)

# Basic setup
setwd("C:/Users/Trevor/Documents/manifest_destiny_the_game/")
players_alph <- c("Dan", "Eli", "John", "Rohin", "Tim", "Trevor")
all_regions <- c("Great Lakes", "International", "Mid-Atlantic",
                 "Mountain/Plains", "New England", "Southeast", "Southwest",
                 "The Frontier")

set_theme(theme_bw())

# Colors
status_colors <- c("Success" = "springgreen3",
                    "Fail" = "firebrick2",
                    "Abandoned/DNA" = "sienna2",
                    "Lost state" = "violet",
                    "Start/Now" = "midnightblue")

all_players <- tibble(player = players_alph)

# Read events from file
events <- read_rds("events.rds")

events2 <- events %>%
    mutate(
        player = list(all_players),
        .after = team
    ) %>%
    select(-arank_last)

current_status <- events2[1, ] %>%
    mutate(

        date = today(),
        team = NA,

        action = "Current Game",
        success = "Start/Now",
        id = NA,
        state = NA,
        region = NA,
        area_sqmi = NA

    )

events3 <- events2 %>%
    add_row(current_status, .before = 1) %>%
    unnest(player)

events4 <- events3 %>%
    pivot_longer(c(matches("^(a?rank|score|states|regions|area|b)_"),
                   -area_sqmi, -states_total),
                 names_to = c("name", "player2"), names_sep = "_") %>%
    pivot_wider() %>%
    filter(
        player == player2
    ) %>%
    group_by(player2) %>%
    mutate(

        state_delta = states - lead(states),

        # These are from the "Game Start" row
        across(c(states, states_total), ~replace_na(.x, 0)),

        success = case_when(
                state_delta == -1 ~ "Lost state",
                action == "Game Start" ~ "Start/Now",
                TRUE ~ success
            )

    ) %>%
    filter(

        # Keep the start/end rows
        # When a state was captured, keep the individuals that were affected
        # If an action was a fail, keep only the row for the person who did it
        action %in% c("Current Game", "Game Start") |
            (success %in% c("Success", "Lost state") & state_delta != 0) |
            (success != "Success" & str_detect(team, player))

    ) %>%
    ungroup() %>%
    select(-player2)

totals <- events4 %>%
    ungroup() %>%
    select(date, states_total) %>%
    mutate(
        delta = states_total - lead(states_total)
    ) %>%
    group_by(date) %>%
    filter(
        delta > 0,
        states_total == max(states_total)
    ) %>%
    rename(
        states = states_total
    ) %>%
    ungroup() %>%
    add_row(
        date = as.Date("2026-02-01"),
        states = 0,
        .after = nrow(.)
    )

```

# Introduction

First, we describe some basic statistics about the game so far.

## State acquisition

```{r basic_stats}

states_inplay <- sort(unique(events4$state[events4$success == "Success"]))
states_attempted <- sort(unique(events4$state))

```

There are 65 total states, 59 in North America now that DC is in play, and six
in the International Region. So far,
`r length(states_inplay)` have been captured and `r length(states_attempted)`
attempted (`r round(100 * length(states_inplay) / length(states_attempted))`%
of attempted states captured). So far,
`r round(100 * length(states_inplay) / 65)`% of the total states have been
captured.

- **Captured**: `r states_inplay`
- **Attempted, but uncaptured**: `r setdiff(states_attempted, states_inplay)`

```{r states, echo=FALSE}

ggplot(events4, aes(x = date, y = states)) +
    geom_step(
        data = totals,
        linewidth = 2.5
    ) +
    geom_step(aes(color = player), linewidth = 2) +
    geom_point(aes(fill = success), shape = 21, size = 3) +
    scale_x_date(date_breaks = "month", date_labels = "%b") +
    scale_y_continuous(sec.axis = dup_axis()) +
    scale_fill_manual(values = status_colors) +
    theme(legend.position = "bottom") +
    labs(x = "Date", y = "Total", title = "States in play", fill = "Status",
         color = "Player") +
    guides(fill = guide_legend(nrow = 2, byrow = TRUE))

state_increase <- lm(states ~ date, data = totals)
states_b <- coef(state_increase)[2]

```

On average, `r round(states_b, 3)` new states were captured each
day, or `r round(30 * states_b, 1)` per month, with an estimated
`r round(states_b * 365, 1)` by the end of the game.

## Region acquisition

We need to calculate the number of unique regions in play (out of 8 total).
The regions that have been captured are:

- `r sort(na.omit(unique(events4$region[events4$success == "Success"])))`

Leaving the following uncaptured regions:

- `r setdiff(all_regions, unique(events4$region[events4$success == "Success"]))`

```{r regions, echo=FALSE}

# On which day was a new region acquired?
regions_acq <- events4 %>%
    filter(
        success %in% c("Success", "Start/Now")
    ) %>%
    select(date, action, region) %>%
    mutate(

        regions_total = map_int(row_number(),
                            ~length(unique(na.omit(region[.x:length(region)]))))
    ) %>%
    distinct() %>%
    mutate(
        delta = regions_total - lead(regions_total, default = -1)
    ) %>%
    group_by(date) %>%
    filter(
        # Account for days where multiple regions were acquired
        action == "Current Game" |
            (regions_total == max(regions_total) & delta > 0)
    ) %>%
    select(-region) %>%
    distinct()

regions <- events4 %>%
    select(date, action, player, regions) %>%
    group_by(player) %>%
    mutate(
        regions = replace_na(regions, 0),
        delta = regions - lead(regions, default = 0)
    ) %>%
    filter(
        delta != 0 | action %in% c("Game Start", "Current Game")
    ) %>%
    group_by(player, date) %>%
    filter(
        regions == max(regions)
    )

ggplot(regions, aes(x = date, y = regions)) +
    geom_step(
        data = regions_acq,
        aes(y = regions_total),
        linewidth = 2.5
    ) +
    geom_step(aes(color = player), linewidth = 2) +
    scale_x_date(date_breaks = "month", date_labels = "%b") +
    scale_y_continuous(breaks = 0:8, limits = c(0, 8), sec.axis = dup_axis()) +
    theme(legend.position = "bottom") +
    labs(x = "Date", y = "Total", title = "States in play", color = "Player") +
    guides(fill = guide_legend(nrow = 2, byrow = TRUE))

```

## Total area captured

Now let's calculate the total area on the board.

```{r area, echo=FALSE}

total_area <- 3536343

area_acq <- events %>%
    filter(
        success == "Success"
    ) %>%
    select(date, state, area_sqmi) %>%
    group_by(state) %>%
    filter(
        # Just get the first time they entered the board
        date == min(date)
    ) %>%
    ungroup() %>%
    mutate(
        area_total = map_int(row_number(), ~sum(area_sqmi[.x:length(area_sqmi)]))
    ) %>%
    group_by(date) %>%
    filter(
        # Account for days where multiple states were acquired
        area_total == max(area_total)
    ) %>%
    select(-state, -area_sqmi) %>%
    distinct() %>%
    ungroup() %>%
    add_row(
        date = today(),
        area_total = max(.$area_total),
        .before = 1
    ) %>%
    add_row(
        date = as.Date("2026-02-01"),
        area_total = 0
    )

area_by_player <- events4 %>%
    select(date, action, success, player, area) %>%
    group_by(player)  %>%
    distinct() %>%
    mutate(
        area = replace_na(area, 0),
        delta = area - lead(area, default = 0)
    ) %>%
    group_by(player, date) %>%
    filter(
        area == max(area)
    )


ggplot(area_by_player, aes(x = date, y = area)) +
    geom_step(
        data = area_acq,
        aes(y = area_total / 1000),
        linewidth = 2.5
    ) +
    geom_step(aes(color = player), linewidth = 2) +
    geom_point(aes(fill = success), shape = 21, size = 3) +
    scale_x_date(date_breaks = "month", date_labels = "%b") +
    scale_y_continuous(
        sec.axis = sec_axis(~ 100 * . / (total_area / 1000),
                            name = "% of total"),
        transform = "sqrt"
    ) +
    scale_fill_manual(values = status_colors) +
    theme(legend.position = "bottom") +
    labs(x = "Date", y = "Sqrt(Total area [x1,000])",
        title = "Area in play") +
    guides(fill = guide_legend(nrow = 2))

```

The above plot's $y$ axis has been square root-transformed to minimize the
effect of claiming Alaska.
So far, we have claimed `r options(scipen=999); max(area_acq$area_total)`
square miles, or `r round(100 *  max(area_acq$area_total) / total_area)`% (of
the US, the rest of the world is assigned no area).

# Time between actions

One goal we have is to offer a bonus to individuals who do not travel for a
while. Here we calculate the time between game events.

```{r intervals, echo=FALSE}

interval <- events %>%
    mutate(
        interval1 = interval(lead(date), date) / days(1),
        .after = date
    ) %>%
    filter_out(
        when_any(
            success == "Lost state",
            action %in% c("Game Start", "Current Game")
        )
    )

interval2 <- events4 %>%
    group_by(player) %>%
    mutate(
        interval2 = interval(lead(date), date) / days(1),
        .after = date
    ) %>%
    filter_out(
        when_any(
            success == "Lost state",
            action %in% c("Game Start", "Current Game")
        )
    )

ggplot(interval, aes(x = interval1, fill = as.ordered(format(date, "%m")))) +
    geom_histogram(binwidth = 1, boundary = 0.5) +
    scale_x_continuous(breaks = seq(0, max(interval$interval1, na.rm = TRUE),
                        by = 7)) +
    labs(x = "Days", y = "Count", title = "Days between actions",
            fill = "Month") +
    theme(legend.position = "bottom")

```

The max time
between all events was `r max(interval$interval1, na.rm = TRUE)` days,
average: `r round(mean(interval$interval1, na.rm = TRUE), 2)`.
But the more relevant metric is by player, where the max per-player time between
events was `r max(interval2$interval2, na.rm = TRUE)` days,
average: `r round(mean(interval2$interval2, na.rm = TRUE), 2)` days.

```{r intervals2, echo = FALSE}

p1 <- ggplot(interval2, aes(x = interval2,
                fill = as.ordered(format(date, "%m")))) +
    geom_histogram(binwidth = 2, boundary = 0.5) +
    scale_x_continuous(breaks = seq(0, max(interval2$interval2, na.rm = TRUE),
                        by = 21)) +
    labs(x = "Days", y = "Count", title = "By player",
            fill = "Month") +
    theme(legend.position = "bottom") +
    facet_wrap(vars(player))

p2 <- ggplot(interval2, aes(x = interval2,
                fill = as.ordered(format(date, "%m")))) +
    geom_histogram(binwidth = 2, boundary = 0.5) +
    scale_x_continuous(breaks = seq(0, max(interval2$interval2, na.rm = TRUE),
                        by = 14)) +
    labs(x = "Days", y = "Count", title = "All",
            fill = "Month") +
    theme(legend.position = "bottom")

p1 + p2 + plot_layout(widths = c(2, 1), guides = "collect") &
    theme(legend.position = "bottom")

```

And the percentiles were as follows:

```{r pctile, echo=FALSE}

qs <- tibble(pctile = seq(0, 1, by = 0.1),
             q1 = quantile(interval2$interval2, probs = seq(0, 1, by = 0.1),
                            na.rm = TRUE))

qs$q1

```

However, this may be skewed by events that took place on the same day
(fails, captures + attacks), so we take out events that happened 0 days after
the previous.

```{r pctile2, echo = FALSE}

qs$q2 <- quantile(interval2$interval2[interval2$interval2 > 0],
            probs = seq(0, 1, by = 0.1), na.rm = TRUE)

qs$q2


```

So if we want drawing a card to be an 90th-percentile action, then that
should be about `r qs$q1[10]` - `r qs$q2[10]` days. The below plot shows
how many days pass between actions at each percentile; i.e. if you want a player
to draw a card between half of all trips, then that should be every
 `r qs$q2[6]` days. This plot excludes intervals of 0 in the calculation of
 the cumulative distribution and goes as high as 5 weeks (35 days).

```{r edcdf, echo = FALSE}

ecdf_no0s <- ecdf(interval2$interval2[interval2$interval2 > 0])

smooth_func <- tibble(days = seq(0, 35)) %>%
    mutate(
        cd = ecdf_no0s(days)
    ) %>%
    filter(
        days > 0
    )

ggplot(smooth_func, aes(y = days, x = cd)) +
    geom_line() +
    geom_point() +
    geom_vline(xintercept = c(0.5, 0.9), color = "red") +
    geom_hline(yintercept = qs$q2[c(6, 10)], color = "red") +
    scale_x_continuous(breaks = seq(0, 1, by = 0.1), limits = c(NA, NA)) +
    labs(x = "Cumulative distribution", y = "Days",
         title = "Cumulative distribution of days between events, per player")

```

# Area bonus

The next question is, is the area bonus ($b$) functioning as a catch-up mechanic,
or it it only reinforcing the leader? The area bonus multipliers, $b^\prime$,
as percent of the
total number of states is as follows, although the last-place player is given
a  $b^\prime$ of 0 in the actual scoring:

```{r ab, echo = FALSE}

b_prime <- round(0.2 * 0.75^(1:6 - 1), 3)
b_prime

b6_20states = 20 * 10 * b_prime[6]

```

These numbers are hard to interpret. At 20 states in play, the smallest
$b^\prime$ ($b^\prime{}_6 = `r b_prime[6]`$) results in a $b$ fairly close to 10
(21 results in a closer $b$, but 20 is nicer):

$$
b_6 = `r b6_20states` = 200 \times b^\prime{}_6
$$

We're going to call $b$ standardized at 20 states $B_{20}$.

```{r ab2, echo = FALSE}

b_prime * 200

```

The below plot shows $B_{20}$ for each game state, colored by date. This is
because early in the game, the leader might have had one or two states and
score an outsize bonus. This highlights that most of the large bonuses with
less than five states came early in the game
This plot does not replace the last-place player with a
score of 0. Player states where they had no area are excluded. Each of these
plots jitters the points to avoid overplotting.

```{r area_bonus, echo=FALSE}

arank <- events4 %>%
    filter(
        !(action %in% c("Game Start", "Current Game")),
        area != 0
    ) %>%
    select(date, states_total, player, states, area, arank) %>%
    mutate(

        b1 = 200 * 0.2 * 0.75^(arank - 1),
        b2 = if_else(arank == 6, 0, b1)

    )

ggplot(arank, aes(x = states, y = b1)) +
    geom_jitter(aes(color = date), width = 0.5, height = 0.5) +
    geom_smooth(method = "lm") +
    scale_y_continuous(limits = c(0, NA)) +
    viridis::scale_color_viridis() +
    theme(legend.position = "bottom")

arank1_lm <- lm(b1 ~ states, data = arank)

```

The $R^2$ for the base model is `r round(summary(arank1_lm)$r.squared, 3)`.

```{r arank_0s, echo = FALSE}

ggplot(arank, aes(x = states, y = b2)) +
    geom_jitter(aes(color = date), width = 0.5, height = 0.5) +
    geom_smooth(method = "lm") +
    viridis::scale_color_viridis() +
    theme(legend.position = "bottom")

arank2_lm <- lm(b2 ~ states, data = arank)

```

The $R^2$ for the with-0s model is `r round(summary(arank2_lm)$r.squared, 3)`.
All players --- except Rohin --- held a state (with area) by Feb. 7,
but that includes Eli, Dan, and
Tim sharing Utah. By March 7, Dan had captured California, setting him out of
that group. Eli and Tim wouldn't be differentiated until April 7, so if we
eliminate events before March 7, we can fit two new models that ignore early
game weirdness.

```{r arank_models}

arank1_mar7_lm <- lm(b1 ~ states, data = filter(arank, date > "2026-03-07"))
arank2_mar7_lm <- lm(b2 ~ states, data = filter(arank, date > "2026-03-07"))

```

The $R^2$s for these models (without and with 0-replacement) are
`r round(summary(arank1_mar7_lm)$r.squared, 3)` and
`r round(summary(arank2_mar7_lm)$r.squared, 3)`, respectively.
These high $R^2$ values suggest that the area bonus is mostly just acting as a
bonus for the leading player.

# Challenges

We set an initial challenge success rate goal of 67%. Naturally, attacks should
have a higher success $(S)$ rate than that. There is some disagreement on how to
calculate success. The purpose of the `Abandoned/DNA` $(A)$ success outcome is
to consider challenges that failed mostly because of out-of-game factors, or
where a final score could not be calculated.

Therefore we have:

- **Success$_1$**: Successes divided by success, fails $(F)$, and DNAs,
    $\sigma_1 = S / (S + F + A)$, i.e., real-world success rate.
- **Success$_2$**: Sucesses divided by successes and fails,
    $\sigma_2 = S / (S + F)$, i.e. success rate
    given that you found time for the challenge.

We have also conventionally been calculating each capture/attack as one event;
regardless of the team size. We can multiply those out, i.e., counting one
for each player, which we'll call $\sigma^*_x$.

The raw counts (collapsed across attack types) are shown below:

```{r successes, echo=FALSE}

challenges <- events %>%
    filter_out(
        success == "Start/Now"
    ) %>%
    mutate(
        action = if_else(str_detect(action, "^Attack"), "Attack", action)
    )

chall_table <- table(challenges$action, challenges$success)

knitr::kable(addmargins(chall_table))

```